---
output: html_document
---
#Peer Assessment 1
Note: Code assumes you have downloaded the Actity Monitoring Dataset found at https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

##Total Number of Steps Taken Per Day:
```{r echo=FALSE, results='hide',include=FALSE}
library(sqldf)
```

```{r message=FALSE}
ActivityData <- read.csv(file = "activity.csv")
ActivityDataProcessed <- na.omit(ActivityData)
StepsPerDay <- sqldf("SELECT date, sum(steps) as 'TotalSteps', median(steps) as 'MedianSteps', avg(steps) as 'MeanSteps' FROM ActivityDataProcessed GROUP BY date ORDER BY date")
hist(StepsPerDay$TotalSteps, main = "Histogram of Total Steps Taken per Day", xlab = "Total Steps", col = "green")
```

Mean and Median Steps Per Day
```{r}
StepsPerDay
```

Mean and Median steps per day across whole study
```{r}
mean(StepsPerDay$TotalSteps)
median(StepsPerDay$TotalSteps)
```

##What is the average daily activity pattern?
```{r message=FALSE}
StepsPerInterval <- sqldf("SELECT interval, sum(steps) as 'TotalSteps', median(steps) as 'MedianSteps', avg(steps) as 'MeanSteps' FROM ActivityDataProcessed GROUP BY interval ORDER BY interval")
plot(x = StepsPerInterval$interval, y = StepsPerInterval$MeanSteps, type = "l", xlab = "Time Interval", ylab = "Mean Number of Steps")
```

Max average steps interval is 835, found below:
```{r}
StepsPerInterval[which.max(StepsPerInterval$MeanSteps), ]

```

##Imputing missing values
```{r}
NumMissingVals <- length(ActivityData$steps) - length(ActivityDataProcessed$steps)
PercentMissing <- (NumMissingVals/length(ActivityData$steps))*100
```
There are `r NumMissingVals` NAs in the Activity Monitoring Dataset, approximately `r PercentMissing`%!

With a simple for loop you can replace the NA values with the mean steps per a given interval:
```{r}
ActivityDataCopy <- ActivityData

for(i in 1:length(ActivityDataCopy$steps)) 
{
  if(is.na(ActivityDataCopy[i,]$steps)) 
    {
      ActivityDataCopy[i,]$steps <- StepsPerInterval[StepsPerInterval$interval == ActivityDataCopy[i,]$interval, ]$MeanSteps 
    } 
}
```
As you can See I have created a copy of the ActivityData data.frame in order to not pollute the original ActivityData.

This, however has no affect on the summary data, as shown below
```{r}
StepsPerDayFromCopyData <- sqldf("SELECT date, avg(steps) as 'MeanStepsWithAddedValues' FROM ActivityDataCopy GROUP BY date ORDER BY date")
AffectOfInput <- merge(StepsPerDay, StepsPerDayFromCopyData, by = "date")
AffectOfInput[c(1,4,5)]
```

##Are there differences in activity patterns between weekdays and weekends?
```{r}
ActivityDataCopy$DayOfWeek <- weekdays(as.POSIXct(ActivityDataCopy$date))
DayOfWeekActivityData <- sqldf("SELECT steps, date, interval, CASE SUBSTR(DayOfWeek, 1,1) WHEN 'S' THEN 'WEEKEND' ELSE 'WEEKDAY' END As 'DayOfWeek' FROM ActivityDataCopy")
DayOfWeekActivityDataSummary <- sqldf("SELECT interval, DayOfWeek, avg(steps) AS 'MeanSteps' FROM DayOfWeekActivityData GROUP BY interval, DayOfWeek ORDER BY interval, DayOfWeek")

plot(x = DayOfWeekActivityDataSummary[DayOfWeekActivityDataSummary$DayOfWeek == "WEEKDAY", ]$interval, y = DayOfWeekActivityDataSummary[DayOfWeekActivityDataSummary$DayOfWeek == "WEEKDAY", ]$MeanSteps, type = "l", xlab = "interval", ylab = "average number of steps", main = "Average Steps Per Interval (WEEKDAY)")

plot(x = DayOfWeekActivityDataSummary[DayOfWeekActivityDataSummary$DayOfWeek == "WEEKEND", ]$interval, y = DayOfWeekActivityDataSummary[DayOfWeekActivityDataSummary$DayOfWeek == "WEEKEND", ]$MeanSteps, type = "l",xlab = "interval", ylab = "average number of steps", main = "Average Steps Per Interval (WEEKEND)")
```

The graphs above show that although peoples days generally started later, they were more active throughout the day.







